% Mit twoside als weitere Option werden die Seitenränder angepasst, sodass beidseitig gedruckt werden kann
\documentclass[pdftex,12pt,a4paper]{report}

\usepackage[]{dmmdbstmpl}
\usepackage{amsmath}
\usepackage{url}
\usepackage{colortbl}

\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator*{\argmin}{arg\,min}

% Boolean - sets document to doublesided.
% Creates empty page between english and german title page as well as english and german abstract
\newboolean{doublesided}
\setboolean{doublesided}{false}
%\setboolean{doublesided}{true}

\global\arbeittyp{Bachelor}

% General information
\global\bearbeiter{Jaanis Fehling}
\global\betreuer{Walid Durani}
\global\aufgabensteller{Prof. Dr. Christian Böhm} % Normally Prof. Böhm
\global\abgabetermin{tt.mm.yyyy}
\global\ort{Munich}

% English information
\global\englTitel{Attention in Mixed-Type Clustering}
\global\englFach{Computer Science}
\global\englAbstracttext{Clustering mixed-type datasets is a difficult task, and many proposed methods do not manage to beat baseline clustering models that were initially designed for continuous data applications. In addition, they often introduce computational complexity, which makes them impractical in real-world applications, were there is large amounts of data present. With autoencoder neural networks gaining popularity, the possibility of using autoencoders in clustering tasks has emerged. The autoencoder can learn a dense representation of the data, which can then be used for the following clustering step. We extend the autoencoder with an attention mechanism, which allow the autoencoder to learn dependencies between features. Further, we extend the attention mechanism to a full Transformer, which is able to provide contextual feature embeddings. We show that these methods can increase clustering performance on some datasets.}


\begin{document}
\languagesselection
\deckblatt
\erklaerung
\inputabstract
\tableofcontents
\input{thesis}
% Abbildungsverzeichnis (kann auch nach dem Inhaltsverzeichnis kommen)
%\listoffigures
% Tabellenverzeichnis (kann auch nach dem Inhaltsverzeichnis kommen)
%\listoftables
\bibliographystyle{plain} 
\bibliography{dmmdbstmpl}         % verwendet dmmdbstmpl.bib
\end{document}
